Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-17 draft.tex
.

Obviously, the appropriate choice of $NO_0$ is crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity, as it was the case for canonical intrinsic homeostasis. Rather, one needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, for practical purposes of the simulation, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}-\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section.
\begin{table}
\begin{tabular}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 12500 $\mathrm{\mu m^2 s^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see section \ref{activ_analys} \\
\hline
\end{tabular}
\caption{Parameters of homeostatic intrinsic plasticity.}
\label{Params_IP}
\end{table}

@@

\section{Discussion}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. As an additional effect, we observed the emergence of a small number of highly influential neurons, which is a feature observed in other theoretical and experimental studies. Section \ref{theor_osc} discussed the emergence of noise-induced oscillations. Suppressing this instability required homeostasis to act on a slow timescale of minutes to hours, which, however, is in accordance with experimentally observed time scales of activity modulation through nitric oxide \cite{Steinert_NO}.

In general, diffusive homeostasis is regarded as a possibility of maintaining an overall level of activity, while allowing single neurons greater variability compared to single-neuron homeostasis, be it a single or individual target rates. According to Sweeney et al., this allows for the encoding of heterogeneous input by differences in the resulting excitatory activity. For this effect to be persistent, one should be able to alter the homeostatic fixed point by changes in external input.
Our results did not support this. As shown in Figure \ref{Ext_Input_Switch}, we found that diffusive homeostasis does not maintain separability of external input by the resulting firing rates. Rather, by analysing network activity and comparing it to our analytic model, we found evidence that the fixed point of network activity is strongly determined and can be well predicted by the spatial configuration of excitatory neurons. However, the validity of this prediction declined for larger diffusion constants and completely vanished for instant diffusion. In a strict sense, only instant diffusion allows for an infinite number of activity configurations that fulfil the homeostatic condition. Presumably however, if diffusion in a real biological system happens at a finite but fast speed compared to the adaptation of intrinsic excitability, one could arguably expect it to be equivalent to instant diffusion. In our simulation, we explained the fact that we saw a steady decline of correlation in Figure \ref{Corr_Coeff_vs_D} by the decreasing capability of the homeostatic mechanism to quickly enough react to structural synaptic changes that alter the recurrent excitatory inputs and, in consequence, excitatory firing rates.

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Of course, this raises the question, whether one can expect such an effect to be observable in the nervous system. To our knowledge, no experimental study exists that directly relates fluctuations of neuronal densities to the existence of highly influential excitatory neurons. Ivenshitz and Segal studied the influence of neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, since our network was tuned to exhibit non-bursting activity, it is hard to draw any conclusions from this result.

>>>>

Obviously, the appropriate choice of $NO_0$ is crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity, as it was the case for canonical intrinsic homeostasis. Rather, one needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, for practical purposes of the simulation, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}-\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section. Diffusion parameters roughly match those measured in experiments \cite{Philippides_2000}.
\begin{table}
\begin{tabular}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 10000 $\mathrm{\mu m^2 s^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see section \ref{activ_analys} \\
\hline
\end{tabular}
\caption{Parameters of homeostatic intrinsic plasticity.}
\label{Params_IP}
\end{table}

@@

\section{Discussion}

\subsection{Homeostatic Stability}
Though being unintended in our recurrent network, the occurrence and necessary analysis of persistent oscillations led to a better understanding of their origin and the need for slow timescales of homeostatic adaptation. In another theoretical study, Harnack et al. treat the necessity of slow homeostatic adaptation in a more general fashion \cite{Stability_Homeostasis_Harnack_2015}. Among other results, it is reported that increasing the number of dynamic variables in a feedback system decreases its stability. This is in line with the fact that our previously used, single-cell homeostatic control (see Equation \eqref{can_hom_rate}) exhibited no instabilities, since it only incorporated firing thresholds as dynamic variables. In contrast, diffusive homeostasis increased the dimensionality of the system by means of its NO-synthesis and diffusion pathway. Additionally Harnack et al. report a similar effect as seen in our system: Noise can reduce stability in a damped dynamical system by constantly causing deviations from its fixed point.

\subsection{The Firing Rate Distribution}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics, we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches to understand the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behaviour of recurrent networks are largely based on the assumption that the excitability of neurons is not individually tuned but rather adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this "canonical" approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal much differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behaviour.

Sweeney et al. claim that their implementation of diffusive homeostasis enables individual neurons to vary their firing rates more flexible compared to single-neuron homeostatic models, especially allowing for sustained encoding of different levels of external input. Interestingly, we could not confirm this, at least within the tests that we performed and the given set of standard parameters. Overall, the resulting behaviour of the excitatory population rather appeared to us as a form of single-neuron homeostasis whose individual targets are predetermined by the quenched random spatial structure of the respective cells. Of course, this impression only applies to the case where the aforementioned mixture of influence between spatial and synaptic topology is dominated by the former. In particular, our result depicted in Figure \ref{Ext_Input_Switch} did not support the notion of separability of inputs. Furthermore, by analyzing network activity and comparing it to our analytic model, we found that only one particular homeostatic fixed point of network activity is exists.

For the standard choice of parameters (being roughly based on experimental data), the resulting activity is indeed strongly bound to the spatial structure of the cells. Naturally, this raises the question of biological plausibility. To our knowledge, no experimental study exists that relates measurements of spontaneous activity to local fluctuations of neuronal density. Ivenshitz and Segal studied the influence of (mean) neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, since our network was tuned to exhibit non-bursting activity, it is hard to draw any conclusions from this result.

From a more general perspective, one could argue that a mechanism binding firing rates to spatial constraints suppresses the ability of the system to adapt to a changing environment since it essentially "freezes" the state of activity. However, experiments have shown that neurons' firing rates are indeed correlated under a changing environment/task: Neurons with a certain baseline-firing rate are very likely to have a similar firing rate while performing a task \cite{Buzsaki_Fir_Rates_2014}. While homeostatic fixation might not be problematic in the case of transient changes of stimuli, a spatial fixation of firing rates would presumably contradict the fact that long-term learning processes and plasticity lead to persistent changes in activity patterns \cite{Lever_Long_Term_Plast_2002}.

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Since this effect is directly linked to the low density-high firing rate-relation shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (C), all previous thoughts regarding plausibility similarly apply to this result: \emph{If} a relationship between density and firing rates existed, we could indeed argue to also expect a relation as shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (D).

  If diffusion in a real biological system happens at a finite but fast speed compared to the adaptation of intrinsic excitability, one could arguably expect it to be equivalent to instant diffusion. In our simulation, we explained the fact that we saw a steady decline of correlation in Figure \ref{Corr_Coeff_vs_D} by the decreasing capability of the homeostatic mechanism to quickly enough react to structural synaptic changes that alter the recurrent excitatory inputs and, in consequence, excitatory firing rates.
